Parallel Tractability of Ontology\\ Materialization: Technique and Practice

### 2018-01-27 parallel-tractability-J-v9/response.tex

\begin{quote}
There should, however, be some discussion somewhere in the paper on the
relationship between path length and KB size.   Even if a KB uses a
non-parallelizable ontology language this doesn't matter much if the paths
that cause problems are short compared to the size of the KB.  The
non-parallel aspects of path processing of short paths will not be noticable
with a small number of threads (certainly up to hundreds) and maybe not even
for very many threads (because of general overhead).
\end{quote}

We added xxx to adress this...

\begin{quote}
  The second sentence of Section 3 is badly stated and misleading.  It is definitely true that for a Datalog program $\langle R, I \rangle$ the rule set $R$ is fixed.  However, the initial facts $I$ are also fixed.
\end{quote}

We made it clear that the rule set is fixed for a class of Datalog programs.
We also revised the similar presentations in Section~2.

@@

\begin{quote}
It is not clear whether the classes of Datalog programs considered in the
paper have to share the same rule set.  The discussion previous to
Definition 1 seems to so indicate, but there is nothing in the definition
that says so.
\end{quote}

We modified Definition~1 to make it clear that the Datalog programs
in a class share the same rule set.

@@

\begin{quote}
There is no definition of what the parents of a node in a graph are.
However, it is quite obvious what the definition should be.
\end{quote}

We added the definition of the parents of a node in a materialization graph in Definition~2.

@@

\begin{quote}
Example 6 does need a bit of revision.  It is not the case that Oex3 cannot
be materialzed tractably in parallel.  Instead there has to be a class of
ontologies that cannot be so materialized.
\end{quote}

We did the revision for Example~6 (see the paragraph following Example~6),
and Example~1 (see the last paragraph of Section~3.2).

@@

\begin{quote}
It would be much better if the ontology names corresponded to the example
numbers, i.e., the ontology in Example 6 was Oex6.
\end{quote}

We made the ontology subscripts correspond to the examples numbers
for Example~5, Example~6, Example~7 and Example~9.

### 2018-01-27 parallel-tractability-J-v9/parallel-tractability-J.tex
fixed quote 5,6 of reviewer 1

\begin{definition}
\textbf{(Materialization Graph)}\label{def:mg}
A \emph{materialization graph}, with respect to
a datalog program $P=\langle R, \textbf{I}\rangle$, is a directed acyclic graph
denoted by $\mathcal{G}=\langle V, E\rangle$ where,
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $V$ is the node set and $V\subseteq T_R^{\omega}(\textbf{I})$;
  \item $E$ is the edge set and $E\subseteq T_R^{\omega}(\textbf{I})\times T_R^{\omega}(\textbf{I})$.
\end{enumerate}
For each edge of the form $e(v_1, v_2)$ where $v_1$ and $v_2$ are the two nodes,
we say that node $v_1$ is the parent of node $v_2$ and node $v_2$ is the child
of node $v_1$. Further, $\mathcal{G}$ satisfies the following condition:
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $\forall H,B_1,...,B_n\in V$ such that $e(B_1, H),...,e(B_n, H)\in E$ and
  $B_1,...,B_n$ are \d{all} the parents of $H$, we have that
  $B_1,...,B_n\rightarrow H\in P^*$.
\end{enumerate}
\end{definition}

@@

The algorithm $\mathsf{A}_{bsc}^{\psi}$ is restricted in the sense that it cannot even work on the
rather simple datalog program $P_{ex_1}$ in Example~\ref{exp:mg}.
The graph $\mathcal{G}_{ex_1}$ in Figure~\ref{fig:mg} is the unique materialization graph of $P_{ex_1}$.
One can also check that \texttt{depth}($\mathcal{G}_{ex_1}$)$=k$.
This means that the depth of $\mathcal{G}_{ex_1}$ is linearly bounded by $k$.
On the other hand, the size of $P_{ex_1}$ is a polynomial of the integer $k$.
Thus, for any $\psi$ that is poly-logarithmically bounded, we can always find a $k$ large
enough such that $\mathsf{A}_{bsc}^{\psi}$ terminates without constructing a materialization
graph of $P_{ex_1}$.
However there indeed exists an \texttt{NC} algorithm that can handle $P_{ex_1}$.
We discuss this in the next part.

@@

In this section, we study the issue of parallel tractability
for materialization of DL-Lite and DHL (DHL($\circ$)) ontologies
based on Algorithm~$\mathsf{A}_{opt}$.
We show that, for any DL-Lite$_{core}$ or DL-Lite$_\mathcal{R}$ ontology $\mathcal{O}$,
there exists a poly-logarithmically bounded function $\psi$
such that Algorithm~$\mathsf{A}_{opt}^{\psi}$ can handle materialization of $\mathcal{O}$.
However, for DHL and DHL($\circ$), there exist ontologies such that Algorithm~$\mathsf{A}_{opt}^\psi$
does not work. We illustrate the reason why Algorithm~$\mathsf{A}_{opt}^\psi$ cannot always
work by studying specific cases.
Further, we propose to restrict the usage of DHL and DHL($\circ$) in order to achieve parallel tractability
of materialization.

@@

\begin{example}\label{exp:dllite}
Given a DL-Lite ontology $\mathcal{O}_{ex_2}$
where its TBox and RBox contain the following axioms:
$A\sqsubseteq B_1$, $A\sqsubseteq B_2$, $B_1\sqcap B_2\sqsubseteq\bot$,
$\exists R.B_2\sqsubseteq\bot$, $Q\sqsubseteq S^-$, $S\sqsubseteq R$;
its ABox contains an assertion $Q(a,b)$.
We denote the corresponding datalog program of $\mathcal{O}_{ex_2}$ by $P_{ex_2}=\langle R, \textbf{I}\rangle$,
where $R$ contains the rules that are transformed from the above axioms; $\textbf{I}$ contains $Q(a,b)$
as the only fact.
The unique materialization graph of $P_{ex_2}$ is denoted by $\mathcal{G}_{ex_2}$ (see Figure~\ref{fig:ex2}).
\end{example}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.6\textwidth]{fig-dllite.eps}
\caption{The materialization graph of $\mathcal{O}_{ex_2}$.}
\label{fig:ex2}
\end{center}
\end{figure}

Consider performing Algorithm~$\mathsf{A}_{opt}$ on the ontology $\mathcal{O}_{ex_2}$ in Example~\ref{exp:dllite}.
(\uppercase\expandafter{\romannumeral1}) First, Algorithm~$\mathsf{A}_{opt}$ adds all ABox assertions (only $Q(a,b)$ here)
to $\mathcal{G}_{ex_2}$ that is initially an empty graph (\ref{alg3:addFacts} of Algorithm~$\mathsf{A}_{opt}$).
(\uppercase\expandafter{\romannumeral2}) In the first iteration of
\ref{alg3:updateG}, Algorithm~$\mathsf{A}_{opt}$ checks that all of the nodes
$A(a)$, $S(b,a)$, $B_1(a)$, $B_2(a)$ and $R(b,a)$ have corresponding SWD paths
starting from $Q(a,b)$. Thus, Algorithm~$\mathsf{A}_{opt}$
adds these nodes to $\mathcal{G}_{ex_2}$ immediately. We take an example of node $B_1(a)$,
for which an SWD path $(Q(a,b),A(a),B_1(a))$ exists.
When updating $\mathcal{G}_{ex_2}$ by adding $B_1(a)$ to it,
Algorithm~$\mathsf{A}_{opt}$ first checks whether the parent node $A(a)$ of $B_1(a)$
has been in $\mathcal{G}_{ex_2}$; if $A(a)$ is already in $\mathcal{G}_{ex_2}$, $B_1(a)$
is added to $\mathcal{G}_{ex_2}$ by creating an edge pointing from $A(a)$ to $B_1(a)$;
if $A(a)$ has not been added to $\mathcal{G}_{ex_2}$, Algorithm~$\mathsf{A}_{opt}$ adds $A(a)$ (resp., $B_1(a)$)
to $\mathcal{G}_{ex_2}$ and creates an edge pointing from $Q(a,b)$ to $A(a)$ (resp.,
an edge pointing from $A(a)$ to $B_1(a)$).\footnote{These two cases may happen simultaneously,
since node $A(a)$ and node $B_1(a)$ are being processed in parallel.} The other nodes
$A(a)$, $S(b,a)$, $B_2(a)$ and $R(b,a)$ are processed similarly.
(\uppercase\expandafter{\romannumeral3})
The two nodes $\bot(a)$ and $\bot(b)$
have no SWD path in the first iteration of \ref{alg3:updateG}.
They are left to be processed in the second iteration.
(\uppercase\expandafter{\romannumeral4})
Finally, Algorithm~$\mathsf{A}_{opt}$ finishes constructing $\mathcal{G}_{ex_2}$
after two iterations (\ref{alg3:halt}).

@@

The above lemma guarantees that all atoms of the form $A(x)$ ($A\neq\bot$) or $R(x,y)$
have to be added to the constructed materialization graph in the first iteration of
\ref{alg3:updateG} by applying Algorithm~$\mathsf{A}_{opt}$.
For each atom of the form $\bot(x)$, there may not exist an SWD path in the first iteration
of \ref{alg3:updateG}, since its derivability depends on its two parent nodes
(see nodes $\bot(a)$ and $\bot(b)$ in $\mathcal{G}_{ex_2}$ of
Example~\ref{exp:dllite}). On the other hand, according to the syntaxes of DL-Lite,
$\bot$ dose not occur on the left hand of any axiom. In other words,
an atom of the form $\bot(x)$ cannot be the parent of any other node.
This allows all atoms of the form $\bot(x)$ being added
to the constructed materialization graph in at most two iterations of
\ref{alg3:updateG} by applying Algorithm~$\mathsf{A}_{opt}$.
Based on the above discussion, for any DL-Lite ontology $\mathcal{O}$,
there always exists a poly-logarithmically bounded function $\psi$ such that
Algorithm~$\mathsf{A}_{opt}^{\psi}$ can handle the materialization of $\mathcal{O}$;
more precisely, we can set that $\psi=2$. This result is consistent with
that in \cite{CalvaneseGLLR07}.
Formally, we use $\mathcal{D}_{\textit{\text{dl\_lite}}}$
to denote the set of all DL-Lite$_{core}$ and DL-Lite$_\mathcal{R}$ ontologies, and give the following theorem.

@@

In this part, we study whether Algorithm~$\mathsf{A}_{opt}^\psi$ can handle DHL ontologies.
Unfortunately there exist DHL ontologies such that Algorithm~$\mathsf{A}_{opt}^\psi$
does not work for any poly-logarithmically bounded function $\psi$.
In the following, we first give such an ontology
to illustrate the reason why Algorithm~$\mathsf{A}_{opt}^\psi$ cannot work.
Based on the analysis of this case, we propose to restrict
the usage of DHL in order to achieve parallel tractability
of materialization.

We find that, an unlimited usage of axioms of the
form $B_1\sqcap B_2\sqsubseteq A$
may make it impossible for Algorithm~$\mathsf{A}_{opt}$ to construct a materialization graph
in a poly-logarithmical number of iterations of \ref{alg3:updateG}.
We use the following example to illustrate it.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{fig-pathtwist.eps}
\caption{(left) The materialization graph of $\mathcal{O}_{ex_3}$;
(right) The materialization graph of $\mathcal{O}_{ex_4}$.}
\label{fig:ex3_4}
\end{center}
\end{figure}

\begin{example}\label{exp:dhl}
Given a DHL ontology $\mathcal{O}_{ex_3}$ where its TBox contains three axioms:
$B_1\sqcap B_2\sqsubseteq A$, $\exists S.A\sqsubseteq B_1$ and $\exists R.A\sqsubseteq B_2$;
the ABox is $\{S(a_i,a_{i-1}), R(a_i,a_{i-1}), A(a_1)\}$
for $2\leq i\leq k$ and $k$ is an integer greater than 2.
We denote the corresponding datalog program of $\mathcal{O}_{ex_3}$ by $P_{ex_3}=\langle R, \textbf{I}\rangle$,
where $R$ contains three rules: `$B_1(x),B_2(x)\rightarrow A(x)$',`$S(x,y),A(y)\rightarrow B_1(x)$' and `$R(x,y),A(y)\rightarrow B_2(x)$'.
The materialization graph of $P_{ex_3}$ constructed by $\mathsf{A}_{opt}$ is denoted by $\mathcal{G}_{ex_3}$ (see Figure~\ref{fig:ex3_4}(left)).
\end{example}

One can check that $\mathcal{G}_{ex_3}$ is the unique materialization graph of $P_{ex_3}$.
Observe that there exists a path (e.g., $A(a_1),B_1(a_2),A(a_2),...,A(a_k)$) between $A(a_1)$ and $A(a_k)$.
When performing Algorithm~$\mathsf{A}_{opt}$ on $P_{ex_3}$, it can be checked that
each node of the form $A(a_i)$ (filled with black color,
and $2\leq i\leq k$) has no SWD path until the $i^{th}$ iteration;
in the $i^{th}$ iteration, the parent nodes of $A(a_i)$ have been added to $\mathcal{G}_{ex_2}$,
and, $A(a_i)$ can also be added to $\mathcal{G}_{ex_2}$.
Thus, Algorithm~$\mathsf{A}_{opt}$ cannot handle $P_{ex_3}$ in a poly-logarithmical
number of iterations.
The intuitive reason is that, at least two paths exist
starting from $A(a_1)$ to $A(a_k)$. These paths \emph{twist} mutually and share the same joint nodes (see the black nodes).
It makes the optimization of acceleration used in Algorithm~$\mathsf{A}_{opt}$ invalid.
That is, for each node $A(a_i)$ ($2\leq i\leq k$), until its parents ($B_1(a_i)$ and $B_2(a_i)$) are added
to $\mathcal{G}_{ex_2}$, there would not exist an available SWD path for $A(a_i)$.
We use `\emph{path twisting}' to represent such cases.

@@

\begin{example}\label{exp:simpleC}
Consider an ontology where the TBox contains three axioms:
$B_1\sqcap B_2\sqsubseteq A$, $\exists S.A\sqsubseteq B_1$ and $B_3\sqsubseteq B_2$;
the ABox is $\{S(a_i,a_{i-1}), B_3(a_i), A(a_1)\}$
for $2\leq i\leq k$ and $k$ is an integer greater than 2.
We denote the corresponding datalog program by $P_{ex_4}$ where the rule set contains:
`$B_1(x),B_2(x)\rightarrow A(x)$',`$S(x,y),A(y)\rightarrow B_1(x)$' and `$B_3(x)\rightarrow B_2(x)$'.
$P_{ex_3}$ has a unique materialization graph denoted by $\mathcal{G}_{ex_4}$ (see Figure~\ref{fig:ex3_4}(right)).
\end{example}

In the above example, for the axiom $B_1\sqcap B_2\sqsubseteq A$, all derived atoms
of the form $B_2(x)$ must not be child nodes of an atom $A(y)$ for some $y$.
This ensures that only one path exists between each two nodes among $A(a_2),...,A(a_k)$.
Further, when constructing $\mathcal{G}_{ex_4}$, Algorithm~$\mathsf{A}_{opt}$ can terminate after
two iterations of \ref{alg3:updateG}. Specifically, in the first iteration, Algorithm~$\mathsf{A}_{opt}$
adds all of the nodes $B_3(a_i)$ and $B_2(a_i)$ ($2\leq i\leq k$) to $\mathcal{G}_{ex_4}$,
since they have corresponding SWD paths;
after that, all the other nodes
can be added to $\mathcal{G}_{ex_4}$ in the second iteration (because each
node has an SWD path).
Motivated by this example, we consider restricting the usage of
the axioms $B_1\sqcap B_2\sqsubseteq A$
such that all atoms of the form $B_1(x)$ or $B_2(x)$ cannot be generated by an atom $A(y)$
for some $y$.
To this end, we first define \emph{simple concepts} as follows:

@@

\begin{example}\label{exp:complexRIA}
Given a DHL$(\circ)$ ontology $\mathcal{O}_{ex_5}$ where its TBox is empty;
the RBox $\mathcal{R}$ contains three axioms:
$R_1\circ R_2\sqsubseteq R$, $R_3\circ R\sqsubseteq R_1$ and $R\circ R_4\sqsubseteq R_2$;
the ABox $\mathcal{A}$ is $\{R(a_1,a_1), R_3(a_i,a_{i-1}), R_4(a_{i-1},a_i)\}$
for $2\leq i\leq k$ and $k$ is an integer greater than 2.
The corresponding datalog program $P_{ex_5}$
contains three rules: `$R_1(x,y),R_2(y,z)\rightarrow R(x,z)$',
`$R_3(x,y),R(y,z)\rightarrow R_1(x,z)$' and `$R(x,y),R_4(y,z)\rightarrow R_2(x,z)$'.
The materialization graph of $P_{ex_5}$ constructed by Algorithm~$\mathsf{A}_{opt}$ is denoted by $\mathcal{G}_{ex_5}$.
\end{example}

One can check that the materialization graph $\mathcal{G}_{ex_5}$ has the same shape as that
of $\mathcal{G}_{ex_3}$ in Figure~\ref{fig:ex3_4}.
A twisted path exists in $\mathcal{G}_{ex_5}$ involving $R(a_i,a_i) (2\leq i\leq k)$ as the joint nodes.
Further, all the roles $R_1$, $R_2$, $R_3$, $R_4$ and $R$ in this example are non-transitive roles.

@@

\begin{example}
For the ontology $\mathcal{O}_{ex_4}$ in Example~\ref{exp:complexRIA}, all of the roles
$R_1$, $R_2$ and $R$ are non-simple roles. Thus, $\mathcal{O}_{ex_4}$ does not follow
the simple-role restriction because of $R_1\circ R_2\sqsubseteq R$.
Consider the ontology $\mathcal{O}_{ex_1}$
in Example~\ref{exp:mg} again. The role $R$ is a non-simple role, while $S$ is a simple role.
Thus $\mathcal{O}_{ex_1}$ follows the simple-role restriction. All the implicit nodes
in $\mathcal{G}_{ex_1}$ have corresponding SWD paths in the first iteration.
Thus, `path twisting' cannot occur when materializing $\mathcal{O}_{ex_1}$ by Algorithm~$\mathsf{A}_{opt}$.
\end{example}

>>>>

\begin{definition}
\textbf{(Materialization Graph)}\label{def:mg}
A \emph{materialization graph}, with respect to
a datalog program $P=\langle R, \textbf{I}\rangle$, is a directed acyclic graph
denoted by $\mathcal{G}=\langle V, E\rangle$ where,
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $V$ is the node set and $V\subseteq T_R^{\omega}(\textbf{I})$;
  \item $E$ is the edge set and $E\subseteq T_R^{\omega}(\textbf{I})\times T_R^{\omega}(\textbf{I})$.
\end{enumerate}
For each edge of the form $e(v_1, v_2)$ where $v_1$ and $v_2$ are the two nodes,
we say that node $v_1$ is the parent node of node $v_2$ and node $v_2$ is the child node
of node $v_1$. Further, $\mathcal{G}$ satisfies the following condition:
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $\forall H,B_1,...,B_n\in V$ such that $e(B_1, H),...,e(B_n, H)\in E$ and
  $B_1,...,B_n$ are \d{all} the parents of $H$, we have that
  $B_1,...,B_n\rightarrow H\in P^*$.
\end{enumerate}
\end{definition}

@@

Consider Example~\ref{exp:mg} again. Let the integer $k$ be a variable. We can
get a class of datalog programs, denoted by $\mathbb{P}_{ex_1}$, where the rule set 
is $\{R(x,y),A(y)\rightarrow A(x), S(x,y),R(y,z)\rightarrow R(x,z)\}$ and the instance
set varies according to $k$. The algorithm $\mathsf{A}_{bsc}^{\psi}$ is restricted in the sense that it cannot even work on the rather simple datalog program class $\mathbb{P}_{ex_1}$.
It can be checked that, for some materialization graph $\mathcal{G}_{ex_1}$
corresponding to a datalog program in $\mathbb{P}_{ex_1}$, \texttt{depth}($\mathcal{G}_{ex_1}$)$=k$ for some $k$.
This means that the depths of the materialization graphs are linearly bounded by $k$.
On the other hand, the sizes of the datalog programs in $\mathbb{P}_{ex_1}$ are polynomial of $k$.
Thus, for any $\psi$ that is poly-logarithmically bounded, we can always find a $k$ large
enough such that $\mathsf{A}_{bsc}^{\psi}$ terminates without constructing a materialization
graph for each datalog program in $\mathbb{P}_{ex_1}$.
However there indeed exists an \texttt{NC} algorithm that can handle $\mathbb{P}_{ex_1}$.
We discuss this in the next part.

@@

In this section, we study the issue of parallel tractability
for materialization of DL-Lite and DHL (DHL($\circ$)) ontologies
based on Algorithm~$\mathsf{A}_{opt}$.
We show that, for any DL-Lite$_{core}$ or DL-Lite$_\mathcal{R}$ ontology $\mathcal{O}$,
there exists a poly-logarithmically bounded function $\psi$
such that Algorithm~$\mathsf{A}_{opt}^{\psi}$ can handle materialization of $\mathcal{O}$.
However, for DHL and DHL($\circ$), there exist ontology classes such that Algorithm~$\mathsf{A}_{opt}^\psi$
does not work. We illustrate the reason why Algorithm~$\mathsf{A}_{opt}^\psi$ cannot always
work by studying specific cases.
Further, we propose to restrict the usage of DHL and DHL($\circ$) in order to achieve parallel tractability
of materialization.

@@

\begin{example}\label{exp:dllite}
Given a DL-Lite ontology $\mathcal{O}_{ex_5}$
where its TBox and RBox contain the following axioms:
$A\sqsubseteq B_1$, $A\sqsubseteq B_2$, $B_1\sqcap B_2\sqsubseteq\bot$,
$\exists R.B_2\sqsubseteq\bot$, $Q\sqsubseteq S^-$, $S\sqsubseteq R$;
its ABox contains an assertion $Q(a,b)$.
We denote the corresponding datalog program of $\mathcal{O}_{ex_5}$ by $P_{ex_5}=\langle R, \textbf{I}\rangle$,
where $R$ contains the rules that are transformed from the above axioms; $\textbf{I}$ contains $Q(a,b)$
as the only fact.
The unique materialization graph of $P_{ex_5}$ is denoted by $\mathcal{G}_{ex_5}$ (see Figure~\ref{fig:ex2}).
\end{example}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.6\textwidth]{fig-dllite.eps}
\caption{The materialization graph of $\mathcal{O}_{ex_5}$.}
\label{fig:ex2}
\end{center}
\end{figure}

Consider performing Algorithm~$\mathsf{A}_{opt}$ on the ontology $\mathcal{O}_{ex_5}$ in Example~\ref{exp:dllite}.
(\uppercase\expandafter{\romannumeral1}) First, Algorithm~$\mathsf{A}_{opt}$ adds all ABox assertions (only $Q(a,b)$ here)
to $\mathcal{G}_{ex_5}$ that is initially an empty graph (\ref{alg3:addFacts} of Algorithm~$\mathsf{A}_{opt}$).
(\uppercase\expandafter{\romannumeral2}) In the first iteration of
\ref{alg3:updateG}, Algorithm~$\mathsf{A}_{opt}$ checks that all of the nodes
$A(a)$, $S(b,a)$, $B_1(a)$, $B_2(a)$ and $R(b,a)$ have corresponding SWD paths
starting from $Q(a,b)$. Thus, Algorithm~$\mathsf{A}_{opt}$
adds these nodes to $\mathcal{G}_{ex_5}$ immediately. We take an example of node $B_1(a)$,
for which an SWD path $(Q(a,b),A(a),B_1(a))$ exists.
When updating $\mathcal{G}_{ex_5}$ by adding $B_1(a)$ to it,
Algorithm~$\mathsf{A}_{opt}$ first checks whether the parent node $A(a)$ of $B_1(a)$
has been in $\mathcal{G}_{ex_5}$; if $A(a)$ is already in $\mathcal{G}_{ex_5}$, $B_1(a)$
is added to $\mathcal{G}_{ex_5}$ by creating an edge pointing from $A(a)$ to $B_1(a)$;
if $A(a)$ has not been added to $\mathcal{G}_{ex_5}$, Algorithm~$\mathsf{A}_{opt}$ adds $A(a)$ (resp., $B_1(a)$)
to $\mathcal{G}_{ex_5}$ and creates an edge pointing from $Q(a,b)$ to $A(a)$ (resp.,
an edge pointing from $A(a)$ to $B_1(a)$).\footnote{These two cases may happen simultaneously,
since node $A(a)$ and node $B_1(a)$ are being processed in parallel.} The other nodes
$A(a)$, $S(b,a)$, $B_2(a)$ and $R(b,a)$ are processed similarly.
(\uppercase\expandafter{\romannumeral3})
The two nodes $\bot(a)$ and $\bot(b)$
have no SWD path in the first iteration of \ref{alg3:updateG}.
They are left to be processed in the second iteration.
(\uppercase\expandafter{\romannumeral4})
Finally, Algorithm~$\mathsf{A}_{opt}$ finishes constructing $\mathcal{G}_{ex_5}$
after two iterations (\ref{alg3:halt}).

@@

The above lemma guarantees that all atoms of the form $A(x)$ ($A\neq\bot$) or $R(x,y)$
have to be added to the constructed materialization graph in the first iteration of
\ref{alg3:updateG} by applying Algorithm~$\mathsf{A}_{opt}$.
For each atom of the form $\bot(x)$, there may not exist an SWD path in the first iteration
of \ref{alg3:updateG}, since its derivability depends on its two parent nodes
(see nodes $\bot(a)$ and $\bot(b)$ in $\mathcal{G}_{ex_5}$ of
Example~\ref{exp:dllite}). On the other hand, according to the syntaxes of DL-Lite,
$\bot$ dose not occur on the left hand of any axiom. In other words,
an atom of the form $\bot(x)$ cannot be the parent of any other node.
This allows all atoms of the form $\bot(x)$ being added
to the constructed materialization graph in at most two iterations of
\ref{alg3:updateG} by applying Algorithm~$\mathsf{A}_{opt}$.
Based on the above discussion, for any DL-Lite ontology $\mathcal{O}$,
there always exists a poly-logarithmically bounded function $\psi$ such that
Algorithm~$\mathsf{A}_{opt}^{\psi}$ can handle the materialization of $\mathcal{O}$;
more precisely, we can set that $\psi=2$. This result is consistent with
that in \cite{CalvaneseGLLR07}.
Formally, we use $\mathcal{D}_{\textit{\text{dl\_lite}}}$
to denote the set of all DL-Lite$_{core}$ and DL-Lite$_\mathcal{R}$ ontologies, and give the following theorem.

@@

In this part, we study whether Algorithm~$\mathsf{A}_{opt}^\psi$ can handle DHL ontologies.
Unfortunately there exist DHL ontology classes such that Algorithm~$\mathsf{A}_{opt}^\psi$
does not work for any poly-logarithmically bounded function $\psi$.
In the following, we first give such a case
to illustrate the reason why Algorithm~$\mathsf{A}_{opt}^\psi$ cannot work.
Based on the analysis of this case, we propose to restrict
the usage of DHL in order to achieve parallel tractability
of materialization.

We find that, an unlimited usage of axioms of the
form $B_1\sqcap B_2\sqsubseteq A$
makes it impossible for Algorithm~$\mathsf{A}_{opt}$ to construct a materialization graph
in a poly-logarithmical number of iterations of \ref{alg3:updateG}.
We use the following example to illustrate it.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{fig-pathtwist.eps}
\caption{(left) The materialization graph of $\mathcal{O}_{ex_6}$;
(right) The materialization graph of $\mathcal{O}_{ex_7}$.}
\label{fig:ex3_4}
\end{center}
\end{figure}

\begin{example}\label{exp:dhl}
Given a DHL ontology $\mathcal{O}_{ex_6}$ where its TBox contains three axioms:
$B_1\sqcap B_2\sqsubseteq A$, $\exists S.A\sqsubseteq B_1$ and $\exists R.A\sqsubseteq B_2$;
the ABox is $\{S(a_i,a_{i-1}), R(a_i,a_{i-1}), A(a_1)\}$
for $2\leq i\leq k$ and $k$ is an integer greater than 2.
We denote the corresponding datalog program of $\mathcal{O}_{ex_6}$ by $P_{ex_6}=\langle R, \textbf{I}\rangle$,
where $R$ contains three rules: `$B_1(x),B_2(x)\rightarrow A(x)$',`$S(x,y),A(y)\rightarrow B_1(x)$' and `$R(x,y),A(y)\rightarrow B_2(x)$'.
The materialization graph of $P_{ex_6}$ constructed by $\mathsf{A}_{opt}$ is denoted by $\mathcal{G}_{ex_6}$ (see Figure~\ref{fig:ex3_4}(left)).
\end{example}

One can check that $\mathcal{G}_{ex_6}$ is the unique materialization graph of $P_{ex_6}$.
Observe that there exists a path (e.g., $A(a_1),B_1(a_2),A(a_2),...,A(a_k)$) between $A(a_1)$ and $A(a_k)$.
When performing Algorithm~$\mathsf{A}_{opt}$ on $P_{ex_6}$, it can be checked that
each node of the form $A(a_i)$ (filled with black color,
and $2\leq i\leq k$) has no SWD path until the $i^{th}$ iteration;
in the $i^{th}$ iteration, the parent nodes of $A(a_i)$ have been added to $\mathcal{G}_{ex_6}$,
and, $A(a_i)$ can also be added to $\mathcal{G}_{ex_6}$.
Similar to the datalog program class $\mathbb{P}_{ex_1}$, we can also get a 
datalog program class $\mathbb{P}_{ex_6}$ for the ontology $\mathcal{O}_{ex_6}$
when $k$ is a variable. Based on the above analysis, Algorithm~$\mathsf{A}_{opt}$ cannot complete the materialization
for the datalog programs of $\mathbb{P}_{ex_6}$ in a poly-logarithmical
number of iterations.
The intuitive reason is that, at least two paths exist
starting from $A(a_1)$ to $A(a_k)$. These paths \emph{twist} mutually and share the same joint nodes (see the black nodes).
It makes the optimization of acceleration used in Algorithm~$\mathsf{A}_{opt}$ invalid.
That is, for each node $A(a_i)$ ($2\leq i\leq k$), until its parents ($B_1(a_i)$ and $B_2(a_i)$) are added
to $\mathcal{G}_{ex_6}$, there would not exist an available SWD path for $A(a_i)$.
We use `\emph{path twisting}' to represent such cases.

@@

\begin{example}\label{exp:simpleC}
Consider an ontology $\mathcal{O}_{ex_7}$ where the TBox contains three axioms:
$B_1\sqcap B_2\sqsubseteq A$, $\exists S.A\sqsubseteq B_1$ and $B_3\sqsubseteq B_2$;
the ABox is $\{S(a_i,a_{i-1}), B_3(a_i), A(a_1)\}$
for $2\leq i\leq k$ and $k$ is an integer greater than 2.
We denote the corresponding datalog program by $P_{ex_7}$ where the rule set contains:
`$B_1(x),B_2(x)\rightarrow A(x)$',`$S(x,y),A(y)\rightarrow B_1(x)$' and `$B_3(x)\rightarrow B_2(x)$'.
$P_{ex_7}$ has a unique materialization graph denoted by $\mathcal{G}_{ex_7}$ (see Figure~\ref{fig:ex3_4}(right)).
\end{example}

In the above example, for the axiom $B_1\sqcap B_2\sqsubseteq A$, all derived atoms
of the form $B_2(x)$ must not be child nodes of an atom $A(y)$ for some $y$.
This ensures that only one path exists between each two nodes among $A(a_2),...,A(a_k)$.
Further, when constructing $\mathcal{G}_{ex_7}$, Algorithm~$\mathsf{A}_{opt}$ can terminate after
two iterations of \ref{alg3:updateG}. Specifically, in the first iteration, Algorithm~$\mathsf{A}_{opt}$
adds all of the nodes $B_3(a_i)$ and $B_2(a_i)$ ($2\leq i\leq k$) to $\mathcal{G}_{ex_7}$,
since they have corresponding SWD paths;
after that, all the other nodes
can be added to $\mathcal{G}_{ex_7}$ in the second iteration (because each
node has an SWD path).
Motivated by this example, we consider restricting the usage of
the axioms $B_1\sqcap B_2\sqsubseteq A$
such that all atoms of the form $B_1(x)$ or $B_2(x)$ cannot be generated by an atom $A(y)$
for some $y$.
To this end, we first define \emph{simple concepts} as follows:

@@

\begin{example}\label{exp:complexRIA}
Given a DHL$(\circ)$ ontology $\mathcal{O}_{ex_9}$ where its TBox is empty;
the RBox $\mathcal{R}$ contains three axioms:
$R_1\circ R_2\sqsubseteq R$, $R_3\circ R\sqsubseteq R_1$ and $R\circ R_4\sqsubseteq R_2$;
the ABox $\mathcal{A}$ is $\{R(a_1,a_1), R_3(a_i,a_{i-1}), R_4(a_{i-1},a_i)\}$
for $2\leq i\leq k$ and $k$ is an integer greater than 2.
The corresponding datalog program $P_{ex_5}$
contains three rules: `$R_1(x,y),R_2(y,z)\rightarrow R(x,z)$',
`$R_3(x,y),R(y,z)\rightarrow R_1(x,z)$' and `$R(x,y),R_4(y,z)\rightarrow R_2(x,z)$'.
The materialization graph of $P_{ex_5}$ constructed by Algorithm~$\mathsf{A}_{opt}$ is denoted by $\mathcal{G}_{ex_9}$.
\end{example}

One can check that the materialization graph $\mathcal{G}_{ex_9}$ has the same shape as that
of $\mathcal{G}_{ex_6}$ in Figure~\ref{fig:ex3_4}.
A twisted path exists in $\mathcal{G}_{ex_9}$ involving $R(a_i,a_i) (2\leq i\leq k)$ as the joint nodes.
Further, all the roles $R_1$, $R_2$, $R_3$, $R_4$ and $R$ in this example are non-transitive roles.

@@

\begin{example}
For the ontology $\mathcal{O}_{ex_7}$ in Example~\ref{exp:complexRIA}, all of the roles
$R_1$, $R_2$ and $R$ are non-simple roles. Thus, $\mathcal{O}_{ex_7}$ does not follow
the simple-role restriction because of $R_1\circ R_2\sqsubseteq R$.
Consider the ontology $\mathcal{O}_{ex_1}$
in Example~\ref{exp:mg} again. The role $R$ is a non-simple role, while $S$ is a simple role.
Thus $\mathcal{O}_{ex_1}$ follows the simple-role restriction. All the implicit nodes
in $\mathcal{G}_{ex_1}$ have corresponding SWD paths in the first iteration.
Thus, `path twisting' cannot occur when materializing $\mathcal{O}_{ex_1}$ by Algorithm~$\mathsf{A}_{opt}$.
\end{example}
